---
title: "Dominance Scoring 2024"
subtitle: "draft" 
author:     | 
    | Quinlan Mann
    | 400420834
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  pdf_document:
    latex_engine: xelatex
    toc: no
    keep_md: TRUE
  word_document:
    toc: no
  html_document:
    toc: no
    keep_md: yes
md_document:
    variant: markdown_github
header-includes:
  - \usepackage{lineno}
  - \linenumbers
  - \usepackage{setspace}
  - \AtBeginEnvironment{tabular}{\singlespacing}
  - \AtBeginEnvironment{lltable}{\singlespacing}
  - \AtBeginEnvironment{tablenotes}{\doublespacing}
linestretch: 1
always_allow_html: true
---

```{r, setup, include=F}
knitr::opts_chunk$set(echo = FALSE, fig.width=6, fig.height=4, warning = F, message = F)
options(digits = 3)
set.seed(19970227)
```

```{r libraries, include=F}
library(tidyverse)
library(reshape2)
library(EloRating)
library(RCurl)
library(kableExtra)
library(knitr)
```

#**Introduction**

#**Methods**

Pukeko were observed at Tawharanui Open Sanctuary in New Zealand (36°22′S, 174°49′E) in the summer 2013 (March-April) and Spring 2024 (August-December). Groups were observed over ten days, prior to the observation period, a stake was placed within the territory, and baited using maize (*Zea mays*) until the group was acclimated to the bait. If a disturbance occurred during the observation period the observation period was abandoned, and an additional observation period was scheduled. The observation period was 30 minutes, and the observer recorded the identity of the dominant and subordinate bird if noticed. The observations were recorded using a Sony HandyCam HDR-CX405 camcorder, and was reviewed a second time to ensure that all interactions were recorded correctly.

The data were analyzed using R (*CITE*) within RStudio (*CITE*). David's scores were calculated using the EloRating package (*CITE*). The steepness of the dominance hierarchies were calculated within groups by taking the slope of the linear regression modelling the normalized David's scores against the within group rank, as described in (De Vries et al. 2006) and bootstrapped P-values were generated by randomizing the results of each interaction between dyads, such that the total number of interactions between dyads remained the same as described in De vries et al. (2006), though the identity of the dominant and subordinate could change.

#**Results**

```{r Datafrmes,include=F, echo=FALSE}
#pulling the data from Github
Doms<-getURL("https://raw.githubusercontent.com/QuinlanMann/Pukeko_Dom_Observations/main/Dom_Obs_Master.csv")
Doms<-read.csv(text=Doms)
#if I'm offline use this line
#Doms<-read.csv("C:\\Users\\quinl\\Desktop\\Grad School\\PUKEKO\\Pukeko_Dom_Observations\\Dom_Obs_Master.csv")

Doms<-subset(Doms, Interaction!= "X")

#summarizing counts of interactions between individual birds within groups
Doms<-Doms%>%
  group_by(Group, Dominant, Subordinate)%>%
  summarize(count=length(Interaction))

#Reassigning each group into unique dataframes
for(H in as.list(unique(Doms$Group))) 
  {assign(paste("DS_",H, sep=""), filter(Doms, Group == H))}

#making a list of dataframes created in the previous step
list_of_objects <- mget(ls(.GlobalEnv, pattern = "DS_*"), envir = .GlobalEnv)

#removing the group identifier column from the list of dfs
list_of_objects <- lapply(list_of_objects, 
                  function(x) x[,-c(1)])

#taking the modified dataframes and writing them to the Global environment
list2env(list_of_objects, .GlobalEnv)
```

```{r Matrix Creation, include=F}
# Function for creating a  matrix to calculate David's Scores
matrix_maker<-function(x){
  # construct 0 matrix of correct dimensions with row and column names
  myMat <- matrix(0, #filled with zeros
                  length(sort(unique(unlist(x[1:2])))), #length of all unique bird name values in rows and columns
                  length(sort(unique(unlist(x[1:2])))), #length of all unique bird name values in rows and columns
                  dimnames = list(sort(unique(unlist(x[1:2]))), #naming rows and columns after all unique bird name values in dataframe
                                  sort(unique(unlist(x[1:2])))))
  # fill in the matrix with matrix indexing on row and column names
  myMat[as.matrix(x[c("Dominant", "Subordinate")])] <- x[["count"]]
  #birds not included in any Dom related interactions are included as interacting with themselves in the dataset, and as such are included as 1 instance of interaction with themselves (impossible), this sets that value to a true zero
  diag(myMat)<-0
  return(myMat)
}

#applying function to the list of objects to create square matrices
list_of_objects<-lapply(list_of_objects, matrix_maker)
list_of_objects2<-list_of_objects

#taking the modified dataframes and writing them to the Global environment
list2env(list_of_objects, .GlobalEnv)
```

```{r David Score Calculations}
#applying ELoRating::DS function to the  
list_of_objects<-lapply(list_of_objects, DS)
#binding all David's Scores into one Dataframe
David_Score<-bind_rows(list_of_objects, .id = "Group")

#viewing all David's scores as a table
kable(David_Score, "latex", booktabs = TRUE, longtable = TRUE, caption = "David Scores within group, calculated usign the EloRating Package (*CITE*), and the Normalized David's scores corrected for chance, as described in De Vries et al. (2006)") %>% 
  collapse_rows(columns = 1)%>%
  kable_styling(latex_options = c("hold_position", "repeat_header"))
```

```{r David Score ranks within groups and plotting to get steepness 1, fig.height=6}
David_Score<-David_Score %>%
  group_by(Group) %>%
  mutate(good_ranks = order(order(normDS, decreasing=TRUE)))

library(ggpmisc)

ggplot(David_Score, aes(x=good_ranks, y=normDS))+
  stat_poly_line() +
  stat_poly_eq(use_label(c("eq", "R2")), size=2) +
  geom_point()+
  facet_wrap(~Group, ncol=4)+
  ggthemes::theme_few()
```

```{r David Score ranks within groups and plotting to get steepness 2, include=F}
ggplot(David_Score, aes(y=normDS, good_ranks, color=Group))+
  geom_point(show.legend = FALSE, aes(shape=Group))+
  geom_smooth(method="lm", show.legend = FALSE)+
  ggthemes::theme_few()

list_of_Scores = split(David_Score, David_Score$Group)

regression_maker = function(data, var1, var2) {
  fit = lm(normDS ~ good_ranks, data = data)
  coef(fit)[2]
  }

slopes<-map_dfr(list_of_Scores, regression_maker, .id = "id")
slopes<-rename(slopes, Slopes="good_ranks")
```

```{r function to randomize matrices and generate a regression, include=F}
reg_randomizer<-function(x){
  #quick check to see if the matrix is square
  x<-as.matrix(x)
  if(length(rownames(x))!=length(colnames(x)))
    return("the matrix is not square")
  #adds the values of the matrix, and transposed matrix together to get
  #the sum of interactions
  x<-x+t(x)
  #removes the lower half so I am not adding the sum twice
  x[lower.tri(x)] <- 0  
  x<-as.data.frame(x)
  #creates the winner ID column so I can make the dataframe long
  x<-tibble::rownames_to_column(x, "BIRD1")
  #unique total value of columns per matrix
  N<-length(colnames(x))
  #creates a unique dataframe that has a list of all interaction pairs (totals)
  x<-pivot_longer(as.data.frame(x), 
                  cols = c(2:N), 
                  names_to = "BIRD2", 
                  values_to = "INT")
  x<-filter(x, INT > 0)
  #creating a random number of wins/losses conditional on how many interactions
  x$WIN<-sapply(x$INT, function(y) sample(0:y, 1))
  x$LOSS<-x$INT-x$WIN
  #creating two dataframes of wins and losses
  wins<-x[,c(1,2,4)]
  wins<-wins%>%
    rename(Dominant=BIRD1, 
           Subordinate=BIRD2, 
           count=WIN)
  loss<-x[,c(2,1,5)]
  loss<-loss%>%
    rename(Dominant=BIRD2, 
           Subordinate=BIRD1, 
           count=LOSS)
  #binding the new dataframes
  x<-rbind(wins,loss)
  
  #making them back to a matrix
  myMat <- matrix(0, #filled with zeros
                  length(sort(unique(unlist(x[1:2])))), #length of all unique bird name values in rows and columns
                  length(sort(unique(unlist(x[1:2])))), #length of all unique bird name values in rows and columns
                  dimnames = list(sort(unique(unlist(x[1:2]))), #naming rows and columns after all unique bird name values in dataframe
                                  sort(unique(unlist(x[1:2])))))
  # fill in the matrix with matrix indexing on row and column names
  myMat[as.matrix(x[c("Dominant", "Subordinate")])] <- x[["count"]]
  
  #DS calculation from EloRating Package
  rn <- rownames(myMat)
  cn <- colnames(myMat)
  tmat <- t(myMat)
  summatrix <- myMat + tmat
  diag(summatrix) <- 0
  summatrix <- replace(summatrix, summatrix == 0, NA)
  l1 <- which(is.na(summatrix), arr.ind = TRUE)
  #this is using the Dij calculation change if you decide to use the Pij calculation
    propmatrix <- (myMat + 0.5)/(tmat + myMat + 1)
  propmatrix <- replace(propmatrix, l1, 0)
  w <- rowSums(propmatrix)
  w2 <- propmatrix %*% w
  l <- rowSums(t(propmatrix))
  l2 <- t(propmatrix) %*% l
  DS <- w + w2 - l - l2
  normDS <- ((DS + ((length(DS)) * (length(DS) - 1))/2))/length(DS)
  good_ranks <- order(order(normDS, decreasing=TRUE))
  res <- data.frame(ID = rn, DS = DS, normDS = normDS, good_ranks = good_ranks)
  res <- res[order(res$DS, decreasing = TRUE), ]
  rownames(res) <- NULL
  #return(res)
  
  #creates linear model and outputs the regression slope
  r2<-summary(lm(normDS ~ good_ranks, data = res))[["coefficients"]][2]
  return(list(c(Slope=r2), Scores=res))
}
```

```{r Randomization and creation of dataset, include=F}
C=200

results <- data.frame(matrix(ncol = length(list_of_objects2), nrow = C))
W <- unique(ls(list_of_objects2))
colnames(results) <- W

for(H in 1:length(list_of_objects2))
{
  values <- list_of_objects2[[H]]
  results[, H] <- as.numeric(replicate(C, 
                                  reg_randomizer(values)[[1]], 
                                  simplify = FALSE))
  
}

S=length(colnames(results))

slopes$p <- sapply(1:S, function(H) {
  mean(results[[H]] < slopes$Slopes[H])
})

results<-tibble::rownames_to_column(results, "iteration")
results<-pivot_longer(results, col=c(2:(length(list_of_objects2)+1)))
results<-rename(results, id=name)
```

```{r Visualization of slopes, fig.height=6}
ggplot(results)+
  geom_density(aes(value), color="darkblue", fill="lightblue")+
  facet_wrap(~ id, ncol=4)+
  geom_vline(data = slopes,
             mapping = aes(xintercept = Slopes), 
             linetype = 'dashed', 
             color="red")+
  ggthemes::theme_few()+
  geom_text(data = slopes, aes(y = 20, x = -.55, 
                               label = ifelse(round(p, 3) == 0, 
                                              "p<0.001", 
                                              paste("p=", round(p, 3), sep=""))))
```

```{r linearity calculations and permutations, include=F}
devries2<-function (m) 
  { diag(m) <- 0
    m <- as.matrix(m)
    total <- m + t(m)
    g <- igraph::graph.adjacency(m, mode = "directed", weighted = TRUE, 
                                 diag = FALSE)
    N <- nrow(m)
    NN <- N * N
    d <- igraph::degree(g, mode = "out")
    hF1 <- (12/((N^3) - N))
    hF2 <- (N - 1)/2
    fixedels <- ((total > 0))
    randomels <- 1 - fixedels
    diag(randomels) <- 0
    fixedvals <- fixedels * (0.5 * (((m > t(m))) - ((t(m) > m)) + 1))
    
    Nperms = 10000
    h0 <- vector(length = Nperms)
    hr <- vector(length = Nperms)
    for (k in 1:Nperms) {
      randmat <- matrix(stats::runif(NN), ncol = N)
      newmat <- fixedvals + randomels * ((randmat > t(randmat)))
      V <- rowSums(newmat)
      h0[k] <- hF1 * sum((V - hF2)^2)
      randmat <- matrix(stats::runif(NN), ncol = N)
      nm <- (randmat > t(randmat))
      Vr <- rowSums(nm)
      hr[k] <- hF1 * sum((Vr - hF2)^2)
    }
    t <- sum((hr >= h0))
    hmod <- mean(h0)
    p <- t/Nperms
    return(list(`h-modified` = hmod, `p-value` = p, `h-rand` = hr))
  }
```

```{r, linearity application and visualization, fig.height=6, echo=F}
linearity<-lapply(list_of_objects2, devries2)
linearity<-bind_rows(linearity, .id = "Group")

ggplot(linearity)+
  geom_histogram(aes(`h-rand`), color="coral4", fill="coral1")+
  facet_wrap(~ Group, ncol=4)+
  geom_vline(data=linearity,
             mapping = aes(xintercept = `h-modified`), 
             linetype = 'dashed', 
             size=1,
             color="black")+
  ggthemes::theme_few()+
  theme(legend.position = "none")+
  geom_text(color="darkblue",aes(y = 2000, x = 0.3,
                label=paste("p=", round(`p-value`, 3), sep="")))
```

