---
title: "Outbred Pūkeko achieve higher dominance scores within groups"
subtitle: "draft" 
author:     | 
    | Quinlan Mann
    | 400420834
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  pdf_document:
    latex_engine: xelatex
    toc: no
    keep_md: FALSE
  word_document:
    toc: no
  html_document:
    toc: no
    keep_md: yes
md_document:
    variant: markdown_github
header-includes:
  - \usepackage{lineno}
  - \linenumbers
  - \usepackage{setspace}
  - \AtBeginEnvironment{tabular}{\singlespacing}
  - \AtBeginEnvironment{lltable}{\singlespacing}
  - \AtBeginEnvironment{tablenotes}{\doublespacing}
linestretch: 1
always_allow_html: true
---

```{r, setup, include=F}
knitr::opts_chunk$set(echo = FALSE, fig.width=6, fig.height=4, warning = F, message = F)
options(digits = 3)
set.seed(19970227)
```

```{r libraries, include=F}
library(tidyverse)
library(reshape2)
library(EloRating)
library(RCurl)
library(kableExtra)
library(knitr)
library(glmmTMB)
library(lme4)
library(lmerTest)
```

# **Abstract**

Inbreeding has been shown to lead to marked decreases in fitness across many metrics, from egg hatching success, immune response, to competitive ability. Pukeko (Porphyrio melanotus melanotus) are a subspecies of rail that joint lays, forms clear breeding groups, and will readily inbreed in parts of their range, likely due to habitat saturation. Access to breeding is limited (especially among females) by their dominance rank within their groups. We therefore hypothesized that outbred birds will have greater reproductive success than inbred birds. To explore the impacts of inbreeding on dominance rank, we therefore predict that we will see outbred birds achieve higher dominance scores than inbred birds. We color banded most birds in these groups and collected a blood sample. We collected 5 hours of video per group and . We used these scores to calculate David’s scores for each bird. We then used PCR to amplify 13 microsatellite loci for each individual bird, and used the fragment sizes to calculate a measure of inbreeding within EMIBD9. We then ran a linear mixed effects model to determine the impacts of inbreeding and sex on dominance scores. We simulated confidence intervals from the standard deviation in our inbreeding coefficient estimates. We see that higher degrees of inbreeding leads to lower dominance scores and ranks. We see that while there is no difference in dominance ranks between males and females, we see that the inbreeding impacts females within a group more strongly than it impacts males. This may have strong implications for group longevity, if more outbred birds are breeding, then that may slow the accumulation of deleterious alleles that often accompanies inbreeding.

# **Keywords**

Inbreeding Depression, Dominance, Joint-laying, cooperative Breeding, Pukeko

\newpage

# **Introduction**

Inbreeding is the mating of two closely related individuals, and it may lead to decreases in fitness (Keller and Waller 2002). While there are cases where inbreeding lead to successful adaptations (i.e. island species/subspecies) through purging and local adaptations (Edmands 1999, Leon-Apodaca et al. 2024) most research has explored the methods wherein individuals avoid inbreeding, either through sex biased dispersal (Lebigre et al 2010), or kin recognition (Ihle & Forstmeier 2013), and how it leads to fitness consequences (Crnokrak & Roff 1999). Due to the relatedness between individuals, the offspring are more likely to receive multiple copies of the same alleles (as the related parent individuals are more likely to share similar alleles) (Hedrick & Garcia-Dorado 2016). This may lead to the expression of recessive alleles, and if these alleles lead to suboptimal phenotypes, the individuals who express the recessive trait will experience reductions in fitness. However, if persistent inbreeding occurs, individuals may accumulate many suboptimal phenotypes (if purging or local adaptation do not occur) (Edmands 1999, Hedrick & Garcia-Dorado 2016, Leon-Apodaca et al. 2024). If inbreeding is widespread in the population, the population may be more likely to become extinct (Biljsma et al. 2000, Hoglund 2009). Supporting these predictions, studies of wild bird populations have shown that inbreeding can lead to decreases in egg viability, chick growth, and may delay the age of first reproduction (Blomqvist et al. 2010, Agudo et al. 2012). Studies have also shown differences in the impacts of inbreeding depression in males and females (CITE), with sexual selection and competitive ability interacting to make inbreeding more costly in some species such as Mice (Mus musculus).

Pūkeko (*Porphyrio melanotus melanotus*) are a species of rail (Rallidae) that form cooperative breeding groups (Craig 1977). Cooperative breeding is a social system where multiple individuals assist in rearing young (Vehrencamp & Quinn 2004). Cooperative breeding can become a preferred breeding strategy in response to many different selective pressures, (i.e. ecological constraints on territories (Craig & Jamieson 1988, Vehrencamp & Quinn 2004), and may result in benefits to breeders such as increased parental care from helpers and co-breeders, or increased indirect fitness through kin selection (Craig & Jamieson 1988, Koenig & Dickinson 2016). In New Zealand, Pūkeko form groups of primarily kin, though there are non-kin within groups (Craig & Jamieson 1988, Vehrencamp & Quinn 2004, Babaei et al. 2025). These kin groups are also often joint laying; a cooperative breeding method where multiple females lay their eggs in a shared nest. The North Island population will often joint lay, forming polygynandrous groups (Jamieson 1997). Groups in the North may therefore experience a high degree of inbreeding, as breeding opportunities are primarily between closely related individuals that remain together on their natal territories (Craig & Jamieson 1988, Vehrencamp & Quinn 2004).

Pūkeko exhibit strong transitive hierarchies (Dey & Quinn 2014), with dominant females monopolizing the reproduction within a group (Jamieson 1997). We also see varying degrees of inbreeding within groups (Babaei et al. 2025), it therefore stands to reason that inbreeding, which can impact behaviour (Franks et al. 2023), size (Hajduk et al. 2016), and competitive ability (Välimäki et al. 2007), could have an impact on dominance hierarchies. We are therefore aiming to explore whether within groups, we see more outbred birds achieve higher dominance scores than inbred birds. We hypothesize that outbred birds will have greater reproductive success than inbred birds. Understanding that reproductive events may be monopolized by more dominant birds (Jamieson 1997), we therefore predict that ee will see outbred birds achieve higher dominance scores than inbred birds.

# **Methods**

## *Field methods:*

Pukeko were observed at Tawharanui Open Sanctuary in New Zealand (36°22′S, 174°49′E) in the summer 2013 (March-April) and Spring 2024 (August-December). Groups were observed over ten days, prior to the observation period, a stake was placed within the territory and baited using maize (*Zea mays*) until the group was acclimated to the bait. If a disturbance occurred during the observation period the observation period was abandoned, and an additional observation period was scheduled. The observation period was 30 minutes, and the observer recorded the identity of the dominant and subordinate bird if noticed. The observations were recorded using a Sony HandyCam HDR-CX405 camcorder, and was reviewed a second time to ensure that all interactions were recorded correctly. The data were analyzed using R (*CITE*) within RStudio (*CITE*). David's scores were calculated using the EloRating package (*CITE*).

## *Lab methods:*

Blood samples were taken during banding; DNA was extracted from these samples using a standard salt extraction procedure (Miller et al. 1988). We amplified 13 microsatellite loci using thermocycling conditions optimized by Babaei et al. (2025). We then took the fragment sizes at all loci for all individuals and calculated inbreeding coefficients (F) using EMIBD9 (Wang 2023). EMIBD9 (wang 2023) also outputs a standard deviation around the calculated F for all individuals.

## *Statistical analyses:*

We took the standard deviation and F and calculated the α and β parameters of a beta distribution for each individual. We then used these parameters to generate a distribution of possible values for each individual, that was sampled 1000 times to generate 1000 simulated datasets. We then ran each dataset through a linear mixed effects model that had David’s score (De Vries et al. 2006) as our response variable, sex and inbreeding coefficient (simulated) as predictor variables, and group as a random effect. We then took the regression coefficient estimates, and standard error from each model and calculated the overall regression coefficient and uncertainty using Rubin’s rule (CITE). We were then able to calculate the p-value and 95% confidence intervals.

We repeated the above process again but replaced David’s score (De Vries et al. 2006) as a response variable with within group rank. Within group rank was the ordered rank based on the David’s score (de Vries et al. 2006), i.e. the highest David’s score (De Vries et al. 2006) was 1, and the lowest David’s score (De Vries et al 2006) was also the highest rank (the total number of birds in the group).

Lastly, we compared the slope of normalized dominance ranks within groups between males and females (i.e. average decrease in dominance between individuals as rank decreases), as described in de Vries et al. (2006). We calculated the normalized David’s score and slopes for each group split between sexes and performed a paired t-test to see if there were differences in slopes between sexes (i.e. were dominance hierarchies more egalitarian in one sex).

# **Results**

```{r Datafrmes, include=F, echo=FALSE}
#pulling the data from Github
Doms<-getURL("https://raw.githubusercontent.com/QuinlanMann/Pukeko_Dom_Observations/main/Dom_Obs_Master.csv")
Doms<-read.csv(text=Doms)
#if I'm offline use this line
#Doms<-read.csv("C:\\Users\\quinl\\Desktop\\Grad School\\PUKEKO\\Pukeko_Dom_Observations\\Dom_Obs_Master.csv")

Doms<-Doms[!Doms$Interaction %in% "X",]
colnames(Doms)[colnames(Doms) == "ï..Group"] <- "Group"

#summarizing counts of interactions between individual birds within groups
Doms<-Doms%>%
  group_by(Group, Dominant, Subordinate)%>%
  summarize(count=length(Interaction))

#Reassigning each group into unique dataframes
for(H in as.list(unique(Doms$Group))) 
  {assign(paste("DS_",H, sep=""), filter(Doms, Group == H))}

#making a list of dataframes created in the previous step
list_of_objects <- mget(ls(.GlobalEnv, pattern = "DS_*"), envir = .GlobalEnv)

#removing the group identifier column from the list of dfs
list_of_objects <- lapply(list_of_objects, 
                  function(x) x[,-c(1)])

#taking the modified dataframes and writing them to the Global environment
list2env(list_of_objects, .GlobalEnv)
```

```{r Matrix Creation, include=F}
# Function for creating a  matrix to calculate David's Scores
matrix_maker<-function(x){
  # construct 0 matrix of correct dimensions with row and column names
  myMat <- matrix(0, #filled with zeros
                  length(sort(unique(unlist(x[1:2])))), #length of all unique bird name values in rows and columns
                  length(sort(unique(unlist(x[1:2])))), #length of all unique bird name values in rows and columns
                  dimnames = list(sort(unique(unlist(x[1:2]))), #naming rows and columns after all unique bird name values in dataframe
                                  sort(unique(unlist(x[1:2])))))
  # fill in the matrix with matrix indexing on row and column names
  myMat[as.matrix(x[c("Dominant", "Subordinate")])] <- x[["count"]]
  #birds not included in any Dom related interactions are included as interacting with themselves in the dataset, and as such are included as 1 instance of interaction with themselves (impossible), this sets that value to a true zero
  diag(myMat)<-0
  return(myMat)
}

#applying function to the list of objects to create square matrices
list_of_objects<-lapply(list_of_objects, matrix_maker)
list_of_objects2<-list_of_objects

#taking the modified dataframes and writing them to the Global environment
list2env(list_of_objects, .GlobalEnv)
```

```{r David Score Calculations, include=F, echo=F}
#applying ELoRating::DS function to the  
list_of_objects<-lapply(list_of_objects, DS)
#binding all David's Scores into one Dataframe
David_Score<-bind_rows(list_of_objects, .id = "Group")

#viewing all David's scores as a table
kable(David_Score, "latex", booktabs = TRUE, longtable = TRUE, caption = "David Scores within group, calculated usign the EloRating Package (*CITE*), and the Normalized David's scores corrected for chance, as described in De Vries et al. (2006)") %>% 
  collapse_rows(columns = 1)%>%
  kable_styling(latex_options = c("hold_position", "repeat_header"))
```

```{r David Score ranks within groups and plotting to get steepness 1, fig.height=6, include=F}
David_Score<-David_Score %>%
  group_by(Group) %>%
  mutate(good_ranks = order(order(normDS, decreasing=TRUE)))

library(ggpmisc)

#making pretty labels
David_Score$Group <- factor(David_Score$Group, levels = c("DS_E_Hay","DS_EAST_CAMP_SE",
                                      "DS_Hay_3W","DS_HAY1_N",
                                      "DS_LookoutSE","DS_NCoastSE",
                                      "DS_NPB NW","DS_NPB_E",
                                      "DS_NPB_S","DS_Rd_Flats_Swede_gate",
                                      "DS_RdFlatAnchorGate",
                                      "DS_ROAD_FLATS_SE","DS_SPB_N_CURVE",
                                      "DS_SPBTrigGate","DS_staff_3",
                                      "DS_SY1_C","DS_SY1_NE",
                                      "DS_SY1_NW","DS_W_camp_C",
                                      "DS_WEST_CAMP_NW"),
                  labels = c("East Hay", "East Camp SE", 
                             "Hay 3W", "Hay 1N", 
                             "Lookout SE", "North Coast SE", 
                             "N. Punchbowl NW", "N. Punchbowl E", 
                             "N. Punchbowl S", "Road Flats: Swede gate",
                             "Road Flats: Parking", 
                             "Road Flats SE", "S. Punchbowl N", 
                             "S. Punchbowl: Trig gate", "Staff 3", 
                             "Stockyard 1 C", "Stockyard 1 NE", 
                             "Stockyard 1 NW", "West Camp C", 
                             "West Camp NW")
)

ggplot(David_Score, aes(x=good_ranks, y=normDS))+
  stat_poly_line() +
  stat_poly_eq(use_label(c("eq")), size=4) +
  geom_point()+
  facet_wrap(~Group, ncol=4)+
  ggthemes::theme_few()
```

```{r David Score ranks within groups and plotting to get steepness 2, include=F}
ggplot(David_Score, aes(y=normDS, good_ranks, color=Group))+
  geom_point(show.legend = FALSE, aes(shape=Group))+
  geom_smooth(method="lm", show.legend = FALSE)+
  ggthemes::theme_few()

list_of_Scores = split(David_Score, David_Score$Group)

regression_maker = function(data, var1, var2) {
  fit = lm(normDS ~ good_ranks, data = data)
  coef(fit)[2]
  }


slopes<-map_dfr(list_of_Scores, regression_maker, .id = "id")
slopes<-rename(slopes, Slopes="good_ranks")
```

```{r function to randomize matrices and generate a regression, include=F}
reg_randomizer<-function(x){
  #quick check to see if the matrix is square
  x<-as.matrix(x)
  if(length(rownames(x))!=length(colnames(x)))
    return("the matrix is not square")
  #adds the values of the matrix, and transposed matrix together to get
  #the sum of interactions
  x<-x+t(x)
  #removes the lower half so I am not adding the sum twice
  x[lower.tri(x)] <- 0  
  x<-as.data.frame(x)
  #creates the winner ID column so I can make the dataframe long
  x<-tibble::rownames_to_column(x, "BIRD1")
  #unique total value of columns per matrix
  N<-length(colnames(x))
  #creates a unique dataframe that has a list of all interaction pairs (totals)
  x<-pivot_longer(as.data.frame(x), 
                  cols = c(2:N), 
                  names_to = "BIRD2", 
                  values_to = "INT")
  x<-filter(x, INT > 0)
  #creating a random number of wins/losses conditional on how many interactions
  x$WIN<-sapply(x$INT, function(y) sample(0:y, 1))
  x$LOSS<-x$INT-x$WIN
  #creating two dataframes of wins and losses
  wins<-x[,c(1,2,4)]
  wins<-wins%>%
    rename(Dominant=BIRD1, 
           Subordinate=BIRD2, 
           count=WIN)
  loss<-x[,c(2,1,5)]
  loss<-loss%>%
    rename(Dominant=BIRD2, 
           Subordinate=BIRD1, 
           count=LOSS)
  #binding the new dataframes
  x<-rbind(wins,loss)
  
  #making them back to a matrix
  myMat <- matrix(0, #filled with zeros
                  length(sort(unique(unlist(x[1:2])))), #length of all unique bird name values in rows and columns
                  length(sort(unique(unlist(x[1:2])))), #length of all unique bird name values in rows and columns
                  dimnames = list(sort(unique(unlist(x[1:2]))), #naming rows and columns after all unique bird name values in dataframe
                                  sort(unique(unlist(x[1:2])))))
  # fill in the matrix with matrix indexing on row and column names
  myMat[as.matrix(x[c("Dominant", "Subordinate")])] <- x[["count"]]
  
  #DS calculation from EloRating Package
  rn <- rownames(myMat)
  cn <- colnames(myMat)
  tmat <- t(myMat)
  summatrix <- myMat + tmat
  diag(summatrix) <- 0
  summatrix <- replace(summatrix, summatrix == 0, NA)
  l1 <- which(is.na(summatrix), arr.ind = TRUE)
  #this is using the Dij calculation change if you decide to use the Pij calculation
    propmatrix <- (myMat + 0.5)/(tmat + myMat + 1)
  propmatrix <- replace(propmatrix, l1, 0)
  w <- rowSums(propmatrix)
  w2 <- propmatrix %*% w
  l <- rowSums(t(propmatrix))
  l2 <- t(propmatrix) %*% l
  DS <- w + w2 - l - l2
  normDS <- ((DS + ((length(DS)) * (length(DS) - 1))/2))/length(DS)
  good_ranks <- order(order(normDS, decreasing=TRUE))
  res <- data.frame(ID = rn, DS = DS, normDS = normDS, good_ranks = good_ranks)
  res <- res[order(res$DS, decreasing = TRUE), ]
  rownames(res) <- NULL
  #return(res)
  
  #creates linear model and outputs the regression slope
  r2<-summary(lm(normDS ~ good_ranks, data = res))[["coefficients"]][2]
  return(list(c(Slope=r2), Scores=res))
}
```

```{r Randomization and creation of dataset, include=F}
C=200

results <- data.frame(matrix(ncol = length(list_of_objects2), nrow = C))
W <- unique(ls(list_of_objects2))
colnames(results) <- W

for(H in 1:length(list_of_objects2))
{
  values <- list_of_objects2[[H]]
  results[, H] <- as.numeric(replicate(C, 
                                  reg_randomizer(values)[[1]], 
                                  simplify = FALSE))
  
}

S=length(colnames(results))

slopes$p <- sapply(1:S, function(H) {
  mean(results[[H]] < slopes$Slopes[H])
})

results<-tibble::rownames_to_column(results, "iteration")
results<-pivot_longer(results, col=c(2:(length(list_of_objects2)+1)))
results<-rename(results, id=name)

results$id<-factor(results$id, levels = c("DS_E_Hay","DS_EAST_CAMP_SE",
                                      "DS_Hay_3W","DS_HAY1_N",
                                      "DS_LookoutSE","DS_NCoastSE",
                                      "DS_NPB NW","DS_NPB_E",
                                      "DS_NPB_S","DS_Rd_Flats_Swede_gate",
                                      "DS_RdFlatAnchorGate",
                                      "DS_ROAD_FLATS_SE","DS_SPB_N_CURVE",
                                      "DS_SPBTrigGate","DS_staff_3",
                                      "DS_SY1_C","DS_SY1_NE",
                                      "DS_SY1_NW","DS_W_camp_C",
                                      "DS_WEST_CAMP_NW"),
                  labels = c("East Hay", "East Camp SE", 
                             "Hay 3W", "Hay 1N", 
                             "Lookout SE", "North Coast SE", 
                             "N. Punchbowl NW", "N. Punchbowl E", 
                             "N. Punchbowl S", "Road Flats: Swede gate",
                             "Road Flats: Parking", 
                             "Road Flats SE", "S. Punchbowl N", 
                             "S. Punchbowl: Trig gate", "Staff 3", 
                             "Stockyard 1 C", "Stockyard 1 NE", 
                             "Stockyard 1 NW", "West Camp C", 
                             "West Camp NW")
)
```

```{r Visualization of slopes, fig.height=6, include=F}
ggplot(results)+
  geom_density(aes(value), color="darkblue", fill="lightblue")+
  facet_wrap(~ id, ncol=4)+
  geom_vline(data = slopes,
             mapping = aes(xintercept = Slopes), 
             linetype = 'dashed', 
             color="red")+
  ggthemes::theme_few()+
  geom_text(data = slopes, aes(y = 20, x = -.55, 
                               label = ifelse(round(p, 3) == 0, 
                                              "p<0.001", 
                                              paste("p=", round(p, 3), sep=""))))
```

```{r linearity calculations and permutations, include=F}
devries2<-function (m) 
  { diag(m) <- 0
    m <- as.matrix(m)
    total <- m + t(m)
    N <- nrow(m)
    NN <- N * N
    hF1 <- (12/((N^3) - N))
    hF2 <- (N - 1)/2
    fixedels <- ((total > 0))
    randomels <- 1 - fixedels
    diag(randomels) <- 0
    fixedvals <- fixedels * (0.5 * (((m > t(m))) - ((t(m) > m)) + 1))
    
    Nperms = 10000
    h0 <- vector(length = Nperms)
    hr <- vector(length = Nperms)
    for (k in 1:Nperms) {
      randmat <- matrix(stats::runif(NN), ncol = N)
      newmat <- fixedvals + randomels * ((randmat > t(randmat)))
      V <- rowSums(newmat)
      h0[k] <- hF1 * sum((V - hF2)^2)
      randmat <- matrix(stats::runif(NN), ncol = N)
      nm <- (randmat > t(randmat))
      Vr <- rowSums(nm)
      hr[k] <- hF1 * sum((Vr - hF2)^2)
    }
    t <- sum((hr >= h0))
    hmod <- mean(h0)
    p <- t/Nperms
    return(list(`h-modified` = hmod, `p-value` = p, `h-rand` = hr))
  }
```

```{r linearity application and visualization, include=F, fig.height=6, echo=F}
linearity<-lapply(list_of_objects2, devries2)
linearity<-bind_rows(linearity, .id = "Group")

linearity$Group<-factor(linearity$Group, levels = c("DS_E_Hay","DS_EAST_CAMP_SE",
                                      "DS_Hay_3W","DS_HAY1_N",
                                      "DS_LookoutSE","DS_NCoastSE",
                                      "DS_NPB NW","DS_NPB_E",
                                      "DS_NPB_S","DS_Rd_Flats_Swede_gate",
                                      "DS_RdFlatAnchorGate",
                                      "DS_ROAD_FLATS_SE","DS_SPB_N_CURVE",
                                      "DS_SPBTrigGate","DS_staff_3",
                                      "DS_SY1_C","DS_SY1_NE",
                                      "DS_SY1_NW","DS_W_camp_C",
                                      "DS_WEST_CAMP_NW"),
                  labels = c("East Hay", "East Camp SE", 
                             "Hay 3W", "Hay 1N", 
                             "Lookout SE", "North Coast SE", 
                             "N. Punchbowl NW", "N. Punchbowl E", 
                             "N. Punchbowl S", "Road Flats: Swede gate",
                             "Road Flats: Parking", 
                             "Road Flats SE", "S. Punchbowl N", 
                             "S. Punchbowl: Trig gate", "Staff 3", 
                             "Stockyard 1 C", "Stockyard 1 NE", 
                             "Stockyard 1 NW", "West Camp C", 
                             "West Camp NW")
)

ggplot(linearity)+
  geom_histogram(aes(`h-rand`), color="coral4", fill="coral1")+
  facet_wrap(~ Group, ncol=4)+
  geom_vline(data=linearity,
             mapping = aes(xintercept = `h-modified`), 
             linetype = 'dashed', 
             size=1,
             color="black")+
  ggthemes::theme_few()+
  theme(legend.position = "none")+
  geom_text(color="darkblue",aes(y = 2000, x = 0.3,
                label=paste("p=", round(`p-value`, 3), sep="")))
```

```{r Beta parameters estimation function, echo=F}
estBeta <- function(mu, sd) {
  var <- sd^2
  alpha <- ((1 - mu) / var - 1 / mu) * mu ^ 2
  beta <- alpha * (1 / mu - 1)
  return(params = list(alpha = alpha, beta = beta))
}
```

```{r fig.cap="David Score and Inbreeding coefficient", fig.height=4, echo=F}
#pulling the data from Github
#Doms<-getURL("https://raw.githubusercontent.com/QuinlanMann/Pukeko_Dom_Observations/main/Dom_Obs_Master.csv")
#Doms<-read.csv(text=Doms)

IDS<-na.omit(read.csv("C:\\Users\\quinl\\Downloads\\ID.csv"))
F_SCORE<-read.csv("E:\\quinl\\Desktop\\Grad School\\PUKEKO\\genetic data\\f_score_prelim.csv")

DF<-merge(David_Score, IDS, by.x = "ID", by.y = "COMBO", all=T, sort=F)
DF<-merge(DF, F_SCORE, by = "PPO")

DF<-DF[rep(seq_len(nrow(DF)), each = 1000),]

for(H in 1:nrow(DF)){
  DF$RAND[H]<-rbeta(1, 
        estBeta(DF$f_mean[H], DF$f_sd[H])[[1]], 
        estBeta(DF$f_mean[H], DF$f_sd[H])[[2]])
  }

DF$ITER<-(c(1:1000))  

slopes2<-subset(DF, SEX!="")%>%
  group_by(ITER)%>%
  summarize(
    int=summary(lmer(DS~RAND*SEX+(1|Group)))[["coefficients"]][[1]],
    fmean=summary(lmer(DS~RAND*SEX+(1|Group)))[["coefficients"]][[2]], 
    SexM=summary(lmer(DS~RAND*SEX+(1|Group)))[["coefficients"]][[3]], 
    fmeanM=summary(lmer(DS~RAND*SEX+(1|Group)))[["coefficients"]][[4]], 
    intse=summary(lmer(DS~RAND*SEX+(1|Group)))[["coefficients"]][[5]],
    fmeanse=summary(lmer(DS~RAND*SEX+(1|Group)))[["coefficients"]][[6]], 
    SexMse=summary(lmer(DS~RAND*SEX+(1|Group)))[["coefficients"]][[7]], 
    fmeanMse=summary(lmer(DS~RAND*SEX+(1|Group)))[["coefficients"]][[8]],
    intdf=summary(lmer(DS~RAND*SEX+(1|Group)))[["coefficients"]][[9]],
    fmeandf=summary(lmer(DS~RAND*SEX+(1|Group)))[["coefficients"]][[10]], 
    SexMdf=summary(lmer(DS~RAND*SEX+(1|Group)))[["coefficients"]][[11]], 
    fmeanMdf=summary(lmer(DS~RAND*SEX+(1|Group)))[["coefficients"]][[12]]
    )

```

```{r Rubins Rule calculations for Davids Score, include=F, echo=F}

#Rubin's Rule for pooling estimates from multiple imputation 
#for inbreeding coefficient
theta_fmean<-mean(slopes2$fmean)
Wbar_fmean<-mean((slopes2$fmeanse)^2)
B_fmean<-(1/(length(slopes2$fmean)-1))*sum((slopes2$fmean-theta_fmean)^2)
T_fmean<-Wbar_fmean+(1+1/length(slopes2$fmean))*B_fmean
se_fmean<-sqrt(T_fmean)
wald_fmean<-theta_fmean^2/se_fmean

qf(p=.05, df1=1, df2=70, lower.tail=FALSE)
wald_fmean
pf(wald_fmean, df1=1, df2=70, lower.tail = FALSE)

#for Sex
theta_SexM<-mean(slopes2$SexM)
Wbar_SexM<-mean((slopes2$SexMse)^2)
B_SexM<-(1/(length(slopes2$SexM)-1))*sum((slopes2$SexM-theta_SexM)^2)
T_SexM<-Wbar_SexM+(1+1/length(slopes2$SexM))*B_SexM
se_SexM<-sqrt(T_SexM)
wald_SexM<-theta_SexM^2/se_SexM

qf(p=.05, df1=1, df2=70, lower.tail=FALSE)
wald_SexM
pf(wald_SexM, df1=1, df2=70, lower.tail = FALSE)

#for inbreeding coefficient * sex
theta_fmeanM<-mean(slopes2$fmeanM)
Wbar_fmeanM<-mean((slopes2$fmeanMse)^2)
B_fmeanM<-(1/(length(slopes2$fmeanM)-1))*sum((slopes2$fmeanM-theta_fmeanM)^2)
T_fmeanM<-Wbar_fmeanM+(1+1/length(slopes2$fmeanM))*B_fmeanM
se_fmeanM<-sqrt(T_fmeanM)
wald_fmeanM<-theta_fmeanM^2/se_fmeanM

wald_fmeanM
pf(wald_fmeanM, df1=1, df2=70, lower.tail = FALSE)

```

David's score was negatively impacted by inbreeding coefficients overall (Fig. 1, `r wald_fmeanM`, `r pf(wald_fmeanM, df1=1, df2=70, lower.tail = FALSE)`). David's score was not impacted by sex (`r wald_SexM`, `r pf(wald_SexM, df1=1, df2=70, lower.tail = FALSE)`), but there was a significant interaction between sex and inbreeding (Fig. 2, `r wald_fmeanM`, `r pf(wald_fmeanM, df1=1, df2=70, lower.tail=F)`).


```{r fig.cap="David's Score v Inbreeding Coeff.", fig.height=4, echo=F}
ggplot(subset(DF, ITER==1 & SEX != ""), aes(f_mean, DS, fill=as.factor(ITER)))+
  geom_jitter()+
  ggthemes::theme_few()+
  geom_smooth(data=subset(DF, SEX != ""), 
              aes(RAND, DS),
              method="lm", 
              color=alpha("grey70", 0.1), se=F, alpha=0.01)+
  geom_smooth(method="lm", color="black", se=F, linetype = "solid")+
  #scale_x_continuous(limits = c(0.01,1))+
  geom_jitter(color="black")+
  theme(legend.position="none")+
  #facet_grid(~SEX)+
  labs(y="David's Score", x="Inbreeding coefficient (F)")
```

```{r Rubins Rule calculations for good_ranks, include=F, echo=F}
slopes3<-subset(DF, SEX!="")%>%
  group_by(ITER)%>%
  summarize(
    int=summary(lmer(good_ranks~RAND*SEX+(1|Group)))[["coefficients"]][[1]],
    fmean=summary(lmer(good_ranks~RAND*SEX+(1|Group)))[["coefficients"]][[2]], 
    SexM=summary(lmer(good_ranks~RAND*SEX+(1|Group)))[["coefficients"]][[3]], 
    fmeanM=summary(lmer(good_ranks~RAND*SEX+(1|Group)))[["coefficients"]][[4]], 
    intse=summary(lmer(good_ranks~RAND*SEX+(1|Group)))[["coefficients"]][[5]],
    fmeanse=summary(lmer(good_ranks~RAND*SEX+(1|Group)))[["coefficients"]][[6]], 
    SexMse=summary(lmer(good_ranks~RAND*SEX+(1|Group)))[["coefficients"]][[7]], 
    fmeanMse=summary(lmer(good_ranks~RAND*SEX+(1|Group)))[["coefficients"]][[8]],
    intdf=summary(lmer(good_ranks~RAND*SEX+(1|Group)))[["coefficients"]][[9]],
    fmeandf=summary(lmer(good_ranks~RAND*SEX+(1|Group)))[["coefficients"]][[10]], 
    SexMdf=summary(lmer(good_ranks~RAND*SEX+(1|Group)))[["coefficients"]][[11]], 
    fmeanMdf=summary(lmer(good_ranks~RAND*SEX+(1|Group)))[["coefficients"]][[12]]
  )

#Rubin's Rule for pooling estimates from multiple imputation 
#for inbreeding coefficient
theta_fmean<-mean(slopes3$fmean)
Wbar_fmean<-mean((slopes3$fmeanse)^2)
B_fmean<-(1/(length(slopes3$fmean)-1))*sum((slopes3$fmean-theta_fmean)^2)
T_fmean<-Wbar_fmean+(1+1/length(slopes3$fmean))*B_fmean
se_fmean<-sqrt(T_fmean)
wald_fmean<-theta_fmean^2/se_fmean

qf(p=.05, df1=1, df2=70, lower.tail=FALSE)
wald_fmean
pf(wald_fmean, df1=1, df2=70, lower.tail = FALSE)

#for Sex
theta_SexM<-mean(slopes3$SexM)
Wbar_SexM<-mean((slopes3$SexMse)^2)
B_SexM<-(1/(length(slopes3$SexM)-1))*sum((slopes3$SexM-theta_SexM)^2)
T_SexM<-Wbar_SexM+(1+1/length(slopes3$SexM))*B_SexM
se_SexM<-sqrt(T_SexM)
wald_SexM<-theta_SexM^2/se_SexM

qf(p=.05, df1=1, df2=70, lower.tail=FALSE)
wald_SexM
pf(wald_SexM, df1=1, df2=70, lower.tail = FALSE)

#for inbreeding coefficient * sex
theta_fmeanM<-mean(slopes3$fmeanM)
Wbar_fmeanM<-mean((slopes3$fmeanMse)^2)
B_fmeanM<-(1/(length(slopes3$fmeanM)-1))*sum((slopes3$fmeanM-theta_fmeanM)^2)
T_fmeanM<-Wbar_fmeanM+(1+1/length(slopes3$fmeanM))*B_fmeanM
se_fmeanM<-sqrt(T_fmeanM)
wald_fmeanM<-theta_fmeanM^2/se_fmeanM

qf(p=.05, df1=1, df2=70, lower.tail=FALSE)
wald_fmeanM
pf(wald_fmeanM, df1=1, df2=70, lower.tail = FALSE)

```

Dominance rank within group was negatively impacted by inbreeding coefficients overall (`r wald_fmeanM`, `r pf(wald_fmeanM, df1=1, df2=70, lower.tail = FALSE)`).Dominance rank within group was not impacted by sex (`r wald_SexM`, `r pf(wald_SexM, df1=1, df2=70, lower.tail = FALSE)`), but there was a non-significant interaction between sex and inbreeding (Fig. 3, `r wald_fmeanM`, `r pf(wald_fmeanM, df1=1, df2=70, lower.tail=F)`).


```{r fig.cap= "David's Score v Inbreeding Coeff separated by sex", fig.height=4, echo=F }
ggplot(subset(DF, ITER==1 & SEX != ""), aes(f_mean, DS, fill=as.factor(ITER)))+
  geom_jitter()+
  ggthemes::theme_few()+
  geom_smooth(data=subset(DF, SEX != ""), 
              aes(RAND, DS),
              method="lm", 
              color=alpha("grey70", 0.1), se=F, alpha=0.01)+
  geom_smooth(method="lm", color="black", se=F, linetype = "solid")+
  #scale_x_continuous(limits = c(0.01,1))+
  geom_jitter(color="black")+
  theme(legend.position="none")+
  facet_grid(~SEX)+
  labs(y="David's Score", x="Inbreeding coefficient (F)")
```


```{r fig.cap="Dominance rank v Inbreeding Coeff separated by sex", fig.height=4, echo=F }
ggplot(subset(DF, ITER==1 & SEX!=""), aes(f_mean, good_ranks, fill=as.factor(ITER)))+
  ggthemes::theme_few()+
  geom_smooth(data=subset(DF, SEX != ""), 
              aes(RAND, good_ranks),
              method="lm", 
              color=alpha("grey70", 0.1), se=F, alpha=0.01)+
  geom_smooth(method="lm", color="black", se=F)+
  #scale_x_continuous(limits = c(0.01,1))+
  geom_point(color="black")+
  theme(legend.position="none")+
  facet_grid(~SEX)+
  labs(y="Within Group Rank", x="Inbreeding coefficient (F)")
```

```{r echo=F}
DF2 <- subset(DF, ITER==1)
DF2 <- droplevels(DF2)
DF2 <- split(DF2, list(DF2$Group,DF2$SEX))
DF2 <- DF2[sapply(DF2, function(x) dim(x)[1]) > 0]
DF2 <- lapply(DF2, function(x) 
  lm(normDS~good_ranks, data = x)[["coefficients"]][[2]])
DF2 <- as.data.frame(DF2)
DF2 <- pivot_longer(DF2, cols = everything())

#define function to extract n characters starting from end
substr_end <- function(x, n){
  substr(x, nchar(x)-n+1, nchar(x))
}

#extract 3 characters starting from end 
DF2$sex <- substr_end(DF2$name, 1)
DF2 <- DF2 %>%
  mutate(GRP = substr(name, 1, nchar(name) - 1))
DF2 <-subset(DF2, sex !=".")
```

```{r fig.cap="Dominance hierarchy slopes by sex"}
ggplot(data=DF2, aes(sex, value, fill=sex))+
  geom_boxplot()+
  geom_line(aes(group=GRP), linetype=2)+
  geom_point()+
  ggthemes::theme_few()+
  labs(y="Dominance Slope", x="Sex", fill="Sex")
```

```{r fig.cap="CHECKING BETAS", include=F, fig.height=6, echo=F}
ggplot(DF, aes(f_mean, RAND))+
  geom_point(alpha=0.05, shape=16)+
  geom_smooth(method="lm", se=T, size=1.5)+
  geom_abline(slope=1, intercept=0, 
              color="red", size=1, 
              linetype="dashed")+
  ggthemes::theme_few()
```

```{r eval=F, echo=F}
Relate<-read.csv("E:\\quinl\\Desktop\\Grad School\\PUKEKO\\relatedness.csv")

Relate<-Relate[,c(1,2,12)]
Relate<-merge(Relate, IDS, by.x = "ID1", by.y="PPO", all = T)
names(Relate)[names(Relate) == 'GRP'] <- 'GRP1'
Relate<-merge(Relate, IDS, by.x = "ID2", by.y="PPO", all = T)
names(Relate)[names(Relate) == 'GRP'] <- 'GRP2'

Relate3<-subset(Relate, GRP1==GRP2 & ID1!=ID2)

Relate3$GRP2 <- factor(Relate3$GRP2, levels = c("ECAMPSE",
                                                          "HAY3W",
                                                          "LOOKOUTSE","NCOASTSE",
                                                          "NPBNW","NPBE",
                                                          "NPBS","RDFLATSSWEDE",
                                                          "RDFLATSANCHOR",
                                                          "RDFLATSSE","SPBCURVE",
                                                          "SPBTRIG","STAFF3",
                                                          "SY1C","SY1NE",
                                                          "SY1NW","WCAMPC",
                                                          "WCAMPNW"),
                            labels = c("East Camp SE", 
                                       "Hay 3W", 
                                       "Lookout SE", "North Coast SE", 
                                       "N. Punchbowl NW", "N. Punchbowl E", 
                                       "N. Punchbowl S", "Road Flats: Swede gate",
                                       "Road Flats: Parking", 
                                       "Road Flats SE", "S. Punchbowl N", 
                                       "S. Punchbowl: Trig gate", "Staff 3", 
                                       "Stockyard 1 C", "Stockyard 1 NE", 
                                       "Stockyard 1 NW", "West Camp C", 
                                       "West Camp NW"))

ggplot(Relate3, aes(GRP1, R_1.2))+
  geom_boxplot()+
  geom_point()+
  ggthemes::theme_few()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+
  labs(x="Breeding groups", y="Kinship Coefficient")

Relate3<-merge(Relate3, slopes, by.x = "GRP2", by.y = "id", all=T)
```

```{r fig.cap="Relatedness and group slope", eval=F, echo=F}
ggplot(Relate3, aes(Slopes, R_1.2))+
  geom_jitter(width=0.0)+
  geom_smooth(method="lm")+
  ggthemes::theme_few()+
  labs(x="Within group dominance hierarchy slopes", y="Kinship Coefficient")
```

```{r Dyadic Dom scores, eval=F, echo=F}

extract_dyads <- function(matrix_data) {
  names <- rownames(matrix_data)
  n <- length(names)
  
  dyads <- data.frame(
    bird1 = character(),
    bird2 = character(),
    wins = numeric(),
    losses = numeric(),
    stringsAsFactors = FALSE
  )
  
  for (i in 1:n) {
    for (j in 1:n) {
      if (i != j) {
        dyads <- rbind(dyads, data.frame(
          bird1 = names[i],
          bird2 = names[j],
          wins = matrix_data[i, j],
          losses = matrix_data[j, i],
          stringsAsFactors = FALSE
        ))
      }
    }
  }
  dyads$Dij <- (dyads$wins / (dyads$wins + dyads$losses))-(((dyads$wins / (dyads$wins + dyads$losses))-0.5)*(1/((dyads$wins + dyads$losses)+1)))
  dyads$Dij<-dyads$Dij
  dyads$Diff <- dyads$Dij - (1-dyads$Dij)
  return(dyads)
}

indiv.dyad<-lapply(list_of_objects2, extract_dyads)

```

```{r eval=F, echo=F}
matrix_maker2<-function(x){
  # construct 0 matrix of correct dimensions with row and column names
  myMat <- matrix(0, #filled with zeros
                  length(sort(unique(unlist(x[1:2])))), #length of all unique bird name values in rows and columns
                  length(sort(unique(unlist(x[1:2])))), #length of all unique bird name values in rows and columns
                  dimnames = list(sort(unique(unlist(x[1:2]))), #naming rows and columns after all unique bird name values in dataframe
                                  sort(unique(unlist(x[1:2])))))
  # fill in the matrix with matrix indexing on row and column names
  myMat[as.matrix(x[c("bird1", "bird2")])] <- x[["Diff"]]
  #birds not included in any Dom related interactions are included as interacting with themselves in the dataset, and as such are included as 1 instance of interaction with themselves (impossible), this sets that value to a true zero
  diag(myMat)<-0
  return(myMat)
}

indiv.dyad<-lapply(indiv.dyad, matrix_maker2)
```

```{r eval=F, echo=F, cleanup}
cleanup<-function(matrix){
  matrix[upper.tri(matrix, diag= T)] <- NA
  x<-as.data.frame(matrix)
  
  x<-as.data.frame(x)
  #creates the winner ID column so I can make the dataframe long
  x<-tibble::rownames_to_column(x, "BIRD1")
  #unique total value of columns per matrix
  N<-length(colnames(x))
  #creates a unique dataframe that has a list of all interaction pairs (totals)
  x<-pivot_longer(as.data.frame(x), 
                  cols = c(2:N), 
                  names_to = "BIRD2", 
                  values_to = "Dyad_Slope")
}

indiv.dyad<-lapply(indiv.dyad, cleanup)
indiv.dyad <- bind_rows(indiv.dyad, .id = "group")

indiv.dyad$Dyad_Slope<-abs(indiv.dyad$Dyad_Slope)*-1
```

```{r eval=F, echo=F}
indiv.dyad$group<-factor(indiv.dyad$group, levels = c("DS_E_Hay","DS_EAST_CAMP_SE",
                                                      "DS_Hay_3W","DS_HAY1_N",
                                                      "DS_LookoutSE","DS_NCoastSE",
                                                      "DS_NPB NW","DS_NPB_E",
                                                      "DS_NPB_S","DS_Rd_Flats_Swede_gate",
                                                      "DS_RdFlatAnchorGate",
                                                      "DS_ROAD_FLATS_SE","DS_SPB_N_CURVE",
                                                      "DS_SPBTrigGate","DS_staff_3",
                                                      "DS_SY1_C","DS_SY1_NE",
                                                      "DS_SY1_NW","DS_W_camp_C",
                                                      "DS_WEST_CAMP_NW"),
                         labels = c("East Hay", "East Camp SE", 
                                    "Hay 3W", "Hay 1N", 
                                    "Lookout SE", "North Coast SE", 
                                    "N. Punchbowl NW", "N. Punchbowl E", 
                                    "N. Punchbowl S", "Road Flats: Swede gate",
                                    "Road Flats: Parking", 
                                    "Road Flats SE", "S. Punchbowl N", 
                                    "S. Punchbowl: Trig gate", "Staff 3", 
                                    "Stockyard 1 C", "Stockyard 1 NE", 
                                    "Stockyard 1 NW", "West Camp C", 
                                    "West Camp NW"))


Relate2<-Relate3[,c(1:5,7)]
names(Relate2)[names(Relate2)=="GRP2"]<-"group"
names(Relate2)[names(Relate2)=="COMBO.x"]<-"BIRD1"
names(Relate2)[names(Relate2)=="COMBO.y"]<-"BIRD2"

Relate2 = merge(indiv.dyad, Relate2, by.x=c("BIRD1", "BIRD2"), by.y=c("BIRD1", "BIRD2"))
Relate2$inv<-abs(Relate2$Dyad_Slope)
Relate2$SZN<-if_else(Relate2$group.y=="East Hay" | Relate2$group.y=="Lookout SE" | Relate2$group.y=="North Coast SE" | Relate2$group.y=="Road Flats: Swede gate" | Relate2$group.y=="Road Flats: Parking" | Relate2$group.y=="S. Punchbowl N" | Relate2$group.y=="S. Punchbowl: Trig gate" | Relate2$group.y=="Staff 3", 
                 "SPring", 
                 "Summer")

Relate2<-Relate2 %>% #removes the unknown groups
  group_by(group.x) %>%
  filter(n() > 5)

ggplot(Relate2, 
       aes(R_1.2, Dyad_Slope))+
  geom_point()+
  geom_smooth(method="lm", se=T)+
  ggthemes::theme_few()+
  theme(legend.position="none")+
  labs(y="Dyadic dominance difference", x="Kinship Coefficient")

relatedness<-glmmTMB(inv~R_1.2+SZN+(1|group.x), 
                     data=Relate2)

performance::check_model(relatedness)

summary(relatedness)
```

```{r eval=F, echo=F}
relatedness<-glmmTMB(inv~R_1.2+SZN+(1|group.x), 
                     data=Relate2)

```
